---
title: "Some BP project insights"
date: "04 Sep 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load necessary packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(gridExtra)
library(knitr)
#library(DescTools)
```


```{r, echo=FALSE}
# Read in data
bpdata <- read.csv("/Users/presidentoliver/Desktop/Practicum in Data Analysis/Project 1 - Blood pressure study/bloodpressurestudy.csv")
#names(bpdata)
```


### Data processing and calculation of different endpoints

```{r}
bpdata <- bpdata %>%
  mutate(subject = factor(subject),
         BPchange = after-before,                                # straight pre/post diff
         rel.BPchange = (after-before)/before,                   # change relative to initial
         BP.pct.drop.120 = 100*(before-after)/(before-120)) %>%  # % drop rel to 120 (interp?)
  rename(ID = subject,
         Drug = drug, 
         SysBP.pre = before, 
         SysBP.post = after)
```


### Mean responses by drug

```{r, echo=FALSE}
bpdata %>%
  group_by(Drug) %>%
  summarize(across(where(is.numeric), mean, .names = "mean_{col}")) %>%
  kable()
```


### SDs of responses by drug

```{r, echo=FALSE}
bpdata %>%
  group_by(Drug) %>%
  summarize(across(where(is.numeric), sd, .names = "sd_{col}")) %>%
  kable()
```


### Visualizations

```{r}
p1 <- ggplot(bpdata) + geom_point(aes(x=Drug, y=BPchange)) + 
  labs(y="Change in Systolic BP")
p2 <- ggplot(bpdata) + geom_point(aes(x=Drug, y=rel.BPchange)) + 
  labs(y="Relative Change in Systolic BP")
p3 <- ggplot(bpdata) + geom_point(aes(x=Drug, y=BP.pct.drop.120)) + 
  labs(y="%age Drop Relative to 120")

grid.arrange(p1, p2, p3, ncol = 2, nrow = 2)
```


### Some Different Modeling Strategies

#### Model 1: One-way ANOVA on BP change
```{r}
# One-way ANOVA to compare BP change among drugs
anova1 <- aov(BPchange ~ Drug, data = bpdata)
summary(anova1)

```


#### Model 2: One-way ANOVA on BP change relative to pre-study BP
```{r}
# One-way ANOVA on relative change in BP
anova2 <- aov(rel.BPchange ~ Drug, data = bpdata)
summary(anova2)

```


#### Model 3: One-way ANOVA on %age Drop Relative to 120
```{r}
# One-way ANOVA on % drop relative to 120 mmHg
anova3 <- aov(BP.pct.drop.120 ~ Drug, data = bpdata)
summary(anova3)


```


#### Model 4: One-way ANCOVA on post-study BP, using pre-study BP as a covariate
```{r, message=FALSE}
# Visualization
ggplot(bpdata, aes(x=SysBP.pre, y=SysBP.post, shape=Drug, colour=Drug, fill=Drug)) +
  geom_smooth(method="lm") +
  geom_point() +
  labs(x = "Pre-study Systolic BP (mmHg)",
       y = "Post-study Systolic BP (mmHg)")

# Modeling

ancova1 <- aov(SysBP.post ~ SysBP.pre + Drug, data = bpdata)
summary(ancova1)

# Perform Tukey test in R
TukeyHSD(ancova1, "Drug")

```


#### Model 5: One-way ANCOVA on change in BP, using pre-study BP as a covariate
```{r, message=FALSE}
# Visualization
ggplot(bpdata, aes(x=SysBP.pre, y=BPchange, shape=Drug, colour=Drug, fill=Drug)) +
  geom_smooth(method="lm") +
  geom_point() +
  labs(x = "Pre-study Systolic BP (mmHg)",
       y = "Change in Systolic BP (mmHg)")

# Modeling
ancova2 <- aov(BPchange ~ SysBP.pre + Drug, data = bpdata)
summary(ancova2)

```


### QUESTIONS
1. How can we compare these models?  Is there a way to do it statistically?
2. What is one overriding concern in all the above models?
3. Are you satisfied that analysis assumptions are being met?  What to do if not?
4. What are some peripheral concerns about the validity of the data?

------

### Assessing the "randomization" claim

Of course we would look at the 'pre' scores to investigate.  If randomization were done, the expectation would be that the distributions of pre-treatment BPs would be the same in all groups.  Approaches to this may include:

1. One way ANOVA (paramteric, strong assumptions made, only compares distributional means)

```{r}
# ANOVA to assess equality of pre-study BP across groups
anova_random <- aov(SysBP.pre ~ Drug, data = bpdata)
summary(anova_random)

```
2. Kruskal-Wallis test (nonparametric one-waky ANOVA on ranks -- assesses location differences only)

```{r}
# Nonparametric test to compare distributions of pre-study BP
kruskal.test(SysBP.pre ~ Drug, data = bpdata)

```
3. Kolmogorov-Smirnov test (general nonparameteric comparison of CDFs)

```{r}
# KS test for comparing distributions (pairwise comparisons can be added)
ks_test_results <- pairwise.t.test(bpdata$SysBP.pre, bpdata$Drug, p.adjust.method = "bonferroni")

ks_test_results

```

